Multilevel generalizations of collective spin dynamics and squeezing

### 2019-10-28 collective_sun/multilevel_spin_notes.tex
simplifying some derivations

Here we derive the effective Hamiltonian induced on the fully
symmetric manifold $\M_\FS$ by the single-particle SOC Hamiltonian
$H_{\t{lat}}^{(\phi)}$ in \eqref{eq:H_lat_SOC_B} in the weak SOC limit
$\abs{\phi}\ll1$.  For brevity, we work in units with the lattice
spacing $a=1$ throughout this section.  Perturbations $V$ to the
interaction Hamiltonian $H_{\t{int}}$ induce the following effective
Hamiltonians on the fully symmetric manifold
$\M_\FS$\cite{bravyi2011schrieffer, perlin2019effective}:
\begin{align}
  H_{\t{eff}}^{(1)} = \P_0 V \P_0,
  &&
  H_{\t{eff}}^{(2)} = -\P_0 V \E V \P_0,
  &&
  \E \equiv \sum_{\Delta>0} \f{\P_\Delta}{\Delta},
  \label{eq:pert}
\end{align}
where $\P_\Delta$ is a projector onto the eigenspace of the
interaction Hamiltonian $H_{\t{int}}$ corresponding to an energy
$\Delta$ above that of fully symmetric states
$\ket{\psi_\FS}\in\M_\FS$.

@@

where $\sum_k \p{e^{isk} - \EE_\ell\sp{e^{is\ell}}} = 0$, so

@@

@@

and in turn

@@

  = -\sum_q \p{e^{isq} - \EE_\ell\sp{e^{is\ell}}}
  D_{K,0}^{(q)} \ket{\psi_\FS}

@@

To evaluate the product $\P_0 \tilde\Z_{Jr} \tilde\Z_{Ks} \P_0$ of
mean-subtracted spin-wave operators projected by $\P_0$ onto the fully
symmetric manifold $\M_\FS$, we expand
\begin{align}
  \P_0 \tilde\Z_{Jr} \tilde\Z_{Ks} \P_0
  = \sum_{p,q} \p{e^{irp}-\EE_k\sp{e^{irk}}}
  \p{e^{isq}-\EE_\ell\sp{e^{is\ell}}}
  \P_0 D_{J,0}^{(p)} D_{K,0}^{(q)} \P_0.
  \label{eq:ZZ_prod_start}
\end{align}
Using the permutational symmetry of the fully symmetric manifold, we
can simplify the $p=q$ terms above to
\begin{align}
  \sum_p \p{e^{irp}-\EE_k\sp{e^{irk}}}
  \p{e^{isp}-\EE_\ell\sp{e^{is\ell}}}
  \P_0 D_{J,0}^{(p)} D_{K,0}^{(p)} \P_0
  = N \cov_{p,q}\sp{e^{irp},e^{isq}}
  \P_0 D_{J,0}^{(0)} D_{K,0}^{(0)} \P_0,
\end{align}
where
\begin{align}
  \cov_{p,q}\sp{X_p,Y_q}
  \equiv \EE_p\sp{\p{X_p-\EE_k\sp{X_k}}\p{Y_p-\EE_\ell\sp{Y_\ell}}}
  = \EE_k\sp{X_k Y_k} - \EE_p\sp{X_p} \EE_q\sp{X_q}
  \label{eq:cov}
\end{align}
is the covariance between $X_p$ and $Y_q$, evaluated over all occupied
quasi-momenta $p,q$.  Similarly, the $p\ne q$ terms in
\eqref{eq:ZZ_prod_start} simplify to
\begin{multline}
  \sum_{p\ne q} \p{e^{irp}-\EE_k\sp{e^{irk}}}
  \p{e^{isq}-\EE_\ell\sp{e^{is\ell}}}
  \P_0 D_{J,0}^{(p)} D_{K,0}^{(q)} \P_0 \\
  = \sum_p \p{e^{irp}-\EE_k\sp{e^{irk}}}
  \sum_{q\ne p} \p{e^{isq}-\EE_\ell\sp{e^{is\ell}}}
  \P_0 D_{J,0}^{(0)} D_{K,0}^{(1)} \P_0,
\end{multline}
where $\sum_q \p{e^{isq}-\EE_\ell\sp{e^{is\ell}}} = 0$, so
\begin{align}
  \sum_{q\ne p} \p{e^{isq}-\EE_\ell\sp{e^{is\ell}}}
  = - \p{e^{isp}-\EE_\ell\sp{e^{is\ell}}},
\end{align}
and in turn
\begin{align}
  \sum_{p\ne q} \p{e^{irp}-\EE_k\sp{e^{irk}}}
  \p{e^{isq}-\EE_\ell\sp{e^{is\ell}}}
  \P_0 D_{J,0}^{(p)} D_{K,0}^{(q)} \P_0
  = - N \cov_{p,q}\sp{e^{irp},e^{isq}}
  \P_0 D_{J,0}^{(0)} D_{K,0}^{(0)} \P_0.
\end{align}
Thus far, we have that
\begin{align}
  \P_0 \tilde\Z_{Jr} \tilde\Z_{Ks} \P_0
  = N \cov_{p,q}\sp{e^{irp},e^{isq}}
  \p{\P_0 D_{J,0}^{(0)} D_{K,0}^{(0)} \P_0
    - \P_0 D_{J,0}^{(0)} D_{K,0}^{(1)} \P_0}.
  \label{eq:ZZ_prod_mid}
\end{align}
In order to write this result in terms of collective operators, we
expand
\begin{multline}
  \P_0 \D_{J,0} \D_{K,0} \P_0
  = \sum_p \P_0 D_{J,0}^{(p)} D_{K,0}^{(p)} \P_0
  + \sum_{p\ne q} \P_0 D_{J,0}^{(p)} D_{K,0}^{(q)} \P_0 \\
  = N \P_0 D_{J,0}^{(0)} D_{K,0}^{(0)} \P_0
  + N \p{N-1} \P_0 D_{J,0}^{(0)} D_{K,0}^{(1)} \P_0,
\end{multline}
which implies that
\begin{align}
  - N \P_0 D_{J,0}^{(0)} D_{K,0}^{(1)} \P_0
  = \f{N}{N-1} \P_0 D_{J,0}^{(0)} D_{K,0}^{(0)} \P_0
  - \f1{N-1} \P_0 \D_{J,0} \D_{K,0} \P_0.
\end{align}
The last ingredient we need to re-write \eqref{eq:ZZ_prod_mid} in
terms of collective operators is
\begin{align}
  N \P_0 D_{J,0}^{(0)} D_{K,0}^{(0)} \P_0
  = N \sum_L f_{J,0;K,0}^{L,0} \P_0 D_{L,0}^{(0)} \P_0
  = \sum_L f_{J,0;K,0}^{L,0} \P_0 \D_{L,0} \P_0,
\end{align}
where $f_{J,0;K,0}^{L,0}$ is a structure constant of the transition
operator algebra, defined in \eqref{eq:trans_prod} of Appendix
\ref{sec:trans_prod}.  Altogether, we then have that
\begin{align}
  N\p{\P_0 D_{J,0}^{(0)} D_{K,0}^{(0)} \P_0
    - \P_0 D_{J,0}^{(0)} D_{K,0}^{(1)} \P_0}
  = \f{N}{N-1} \sum_L f_{J,0;K,0}^{L,0} \P_0 \D_{L,0} \P_0
  - \f1{N-1} \P_0 \D_{J,0} \D_{K,0} \P_0,
  \label{eq:eff_sym_drive}
\end{align}
which, substituted into \eqref{eq:ZZ_prod_mid}, implies that
\begin{align}
  \P_0 \tilde\Z_{Jr} \tilde\Z_{Ks} \P_0
  = -\f{\cov_{p,q}\sp{e^{irp},e^{isq}}}{N-1}
  \p{\P_0 \D_{J,0} \D_{K,0} \P_0
    - N \sum_L f_{J,0;K,0}^{L,0} \P_0 \D_{L,0} \P_0}.
  \label{eq:ZZ_prod}
\end{align}

@@

The effective Hamiltonians in \eqref{eq:H_eff_1} and
\eqref{eq:H_eff_2} are respectively first and second order in the SOC
Hamiltonian $H_{\t{lat}}^{(\phi)}$.  Here we extract the leading and
next-to-leading order dependence of these effective Hamiltonians on
the SOC angle $\phi$ using the expansion
\begin{align}
  H_{\t{eff}}^{(1)}
  &= \phi t \, \EE_p\sp{\sin\p{pa}} \xi_1 \D_{1,0}
  + \phi^2 t \, \EE_q\sp{\cos\p{qa}} \xi_2 \D_{2,0}
  + O\p{\phi^3},
  \label{eq:H_eff_1_phi_apndx} \\
  H_{\t{eff}}^{(2)}
  &= \f{\phi^2 t^2}{\p{N-1}fU} \var_q\sp{\sin\p{qa}}
  \p{\xi_1^2 \D_{1,0}^2 - 2 N \xi_2 \D_{2,0}}
  + O\p{\phi^3},
  \label{eq:H_eff_2_phi_apndx}
\end{align}
where the factors $\xi_L$ are defined in \eqref{eq:scale_fac},
\begin{align}
  \var_q\sp{X_q} \equiv \cov_{p,q}\sp{X_p,X_q}
  = \EE_p\sp{\p{X_p-\EE_q\sp{X_q}}^2}
  = \EE_p\sp{X_p^2} - \EE_q\sp{X_q}^2
\end{align}
is the variance of $X_q$ over all occupied quasi-momenta $q$, and we
have neglected scalar terms $\sim\D_{0,0}\propto\1$ that have no
physical consequence.  Note that in the case of SU(2), the drive
operator $\tilde\D_{1,0}$ is precisely the collective spin-$z$
operator $S_\z$ that appears in Ref.~\cite{he2019engineering}.
Furthermore, drive operators $D_{LM}$ vanish when $L\ge n$, so in the
case of SU(2) the operator $\tilde\D_{2,0}$ has no contribution to the
effective Hamiltonians in \eqref{eq:H_eff_2_phi_apndx} and
\eqref{eq:H_eff_2_phi_apndx}.

>>>>

Here we derive the effective Hamiltonian induced on the fully
symmetric manifold $\M_\FS$ by the single-particle SOC Hamiltonian
$H_{\t{lat}}^{(\phi)}$ in \eqref{eq:H_lat_SOC_B} in the weak SOC limit
$\abs{\phi}\ll1$.  For brevity, we work in units with the lattice
spacing $a=1$ throughout this section.  Perturbations $V$ to the
interaction Hamiltonian $H_{\t{int}}$ induce the following effective
Hamiltonians on the fully symmetric manifold
$\M_\FS$\cite{bravyi2011schrieffer, perlin2019effective}:
\begin{align}
  H_{\t{eff}}^{(1)} = \P_0 V \P_0,
  &&
  H_{\t{eff}}^{(2)} = -\P_0 V \E V \P_0,
  &&
  \E \equiv \sum_{k>0} \f{\P_k}{\Delta_k},
  \label{eq:pert}
\end{align}
where $\P_k$ is a projector onto the $k$-th excited eigenspace of the
interaction Hamiltonian $H_{\t{int}}$, with interaction energy
$\Delta_k$ above that of fully symmetric states
$\ket{\psi_\FS}\in\M_\FS$.

@@

where the sum over $k\ne q$ is equal to a sum over all $k$ minus a
term with $k=q$, i.e.

@@

  = \sum_k \p{e^{isk} - \EE_\ell\sp{e^{is\ell}}}
  - \p{e^{isq} - \EE_\ell\sp{e^{is\ell}}}

@@

so

@@

@@

To evaluate the product $\P_0 \tilde\Z_{Jr} \tilde\Z_{Ks} \P_0$ of
mean-subtracted spin-wave operators projected by $\P_0$ onto the fully
symmetric manifold $\M_\FS$, we expand
\begin{align}
  \P_0 \tilde\Z_{Jr} \tilde\Z_{Ks} \P_0
  = \sum_{p,q} \p{e^{irp}-\EE_k\sp{e^{irk}}}
  \p{e^{isq}-\EE_\ell\sp{e^{is\ell}}}
  \P_0 D_{J,0}^{(p)} D_{K,0}^{(q)} \P_0.
  \label{eq:ZZ_prod_start}
\end{align}
Using the permutational symmetry of the fully symmetric manifold, we
can simplify the $p=q$ terms above to
\begin{align}
  \sum_p \p{e^{irp}-\EE_k\sp{e^{irk}}}
  \p{e^{isp}-\EE_\ell\sp{e^{is\ell}}}
  \P_0 D_{J,0}^{(p)} D_{K,0}^{(p)} \P_0
  = N \cov_{p,q}\sp{e^{irp},e^{isq}}
  \P_0 D_{J,0}^{(0)} D_{K,0}^{(0)} \P_0,
\end{align}
where
\begin{align}
  \cov_{p,q}\sp{X_p,Y_q}
  \equiv \EE_p\sp{\p{X_p-\EE_k\sp{X_k}}\p{Y_p-\EE_\ell\sp{Y_\ell}}}
  = \EE_k\sp{X_k Y_k} - \EE_p\sp{X_p} \EE_q\sp{X_q}
  \label{eq:cov}
\end{align}
is the covariance between $X_p$ and $Y_q$, evaluated over all occupied
quasi-momenta $p,q$.  Similarly, the $p\ne q$ terms in
\eqref{eq:ZZ_prod_start} simplify to
\begin{multline}
  \sum_{p\ne q} \p{e^{irp}-\EE_k\sp{e^{irk}}}
  \p{e^{isq}-\EE_\ell\sp{e^{is\ell}}}
  \P_0 D_{J,0}^{(p)} D_{K,0}^{(q)} \P_0 \\
  = \sum_p \p{e^{irp}-\EE_k\sp{e^{irk}}}
  \sum_{q\ne p} \p{e^{isq}-\EE_\ell\sp{e^{is\ell}}}
  \P_0 D_{J,0}^{(0)} D_{K,0}^{(1)} \P_0,
\end{multline}
where the sum over $q\ne p$ is equal to a sum over all $q$ minus a
term with $q=p$, i.e.
\begin{align}
  \sum_{q\ne p} \p{e^{isq}-\EE_\ell\sp{e^{is\ell}}}
  = \sum_q \p{e^{isq}-\EE_\ell\sp{e^{is\ell}}}
  - \p{e^{isp}-\EE_\ell\sp{e^{is\ell}}}
  = - \p{e^{isp}-\EE_\ell\sp{e^{is\ell}}},
\end{align}
which implies that the overall sum over $p,q$ is
\begin{align}
  -\sum_p \p{e^{irp}-\EE_k\sp{e^{irk}}}
  \p{e^{isp}-\EE_\ell\sp{e^{is\ell}}}
  = - N \cov_{p,q}\sp{e^{irp},e^{isq}}.
\end{align}
Thus far, we have that
\begin{align}
  \P_0 \tilde\Z_{Jr} \tilde\Z_{Ks} \P_0
  = N \cov_{p,q}\sp{e^{irp},e^{isq}}
  \p{\P_0 D_{J,0}^{(0)} D_{K,0}^{(0)} \P_0
    - \P_0 D_{J,0}^{(0)} D_{K,0}^{(1)} \P_0}.
  \label{eq:ZZ_prod_mid}
\end{align}
In order to write this result in terms of collective operators, we
expand
\begin{multline}
  \P_0 \D_{J,0} \D_{K,0} \P_0
  = \sum_p \P_0 D_{J,0}^{(p)} D_{K,0}^{(p)} \P_0
  + \sum_{p\ne q} \P_0 D_{J,0}^{(p)} D_{K,0}^{(q)} \P_0 \\
  = N \P_0 D_{J,0}^{(0)} D_{K,0}^{(0)} \P_0
  + N \p{N-1} \P_0 D_{J,0}^{(0)} D_{K,0}^{(1)} \P_0,
\end{multline}
which implies that
\begin{align}
  - N \P_0 D_{J,0}^{(0)} D_{K,0}^{(1)} \P_0
  = \f{N}{N-1} \P_0 D_{J,0}^{(0)} D_{K,0}^{(0)} \P_0
  - \f1{N-1} \P_0 \D_{J,0} \D_{K,0} \P_0.
\end{align}
The last ingredient we need to re-write \eqref{eq:ZZ_prod_mid} in
terms of collective operators is
\begin{align}
  N \P_0 D_{J,0}^{(0)} D_{K,0}^{(0)} \P_0
  = N \sum_L f_{J,0;K,0}^{L,0} \P_0 D_{L,0}^{(0)} \P_0
  = \sum_L f_{J,0;K,0}^{L,0} \P_0 \D_{L,0} \P_0,
\end{align}
where $f_{J,0;K,0}^{L,0}$ is a structure constant of the transition
operator algebra, defined in \eqref{eq:trans_prod} of Appendix
\ref{sec:trans_prod}.  Altogether, we then have that
\begin{align}
  N\p{\P_0 D_{J,0}^{(0)} D_{K,0}^{(0)} \P_0
    - \P_0 D_{J,0}^{(0)} D_{K,0}^{(1)} \P_0}
  = \f{N}{N-1} \sum_L f_{J,0;K,0}^{L,0} \P_0 \D_{L,0} \P_0
  - \f1{N-1} \P_0 \D_{J,0} \D_{K,0} \P_0,
  \label{eq:eff_sym_drive}
\end{align}
which, substituted into \eqref{eq:ZZ_prod_mid}, implies that
\begin{align}
  \P_0 \tilde\Z_{Jr} \tilde\Z_{Ks} \P_0
  = -\f{\cov_{p,q}\sp{e^{irp},e^{isq}}}{N-1}
  \p{\P_0 \D_{J,0} \D_{K,0} \P_0
    - N \sum_L f_{J,0;K,0}^{L,0} \P_0 \D_{L,0} \P_0}.
  \label{eq:ZZ_prod}
\end{align}

@@

The effective Hamiltonians in \eqref{eq:H_eff_1} and
\eqref{eq:H_eff_2} are respectively first and second order in the SOC
Hamiltonian $H_{\t{lat}}^{(\phi)}$.  Here we extract the leading and
next-to-leading order dependence of these effective Hamiltonians on
the SOC angle $\phi$ using the expansion
\begin{align}
  H_{\t{eff}}^{(1)}
  &= \phi t \, \EE_p\sp{\sin\p{pa}} \xi_1 \D_{1,0}
  + \phi^2 t \, \EE_q\sp{\cos\p{qa}} \xi_2 \D_{2,0}
  + O\p{\phi^3},
  \label{eq:H_eff_1_phi_apndx} \\
  H_{\t{eff}}^{(2)}
  &= \f{\phi^2 t^2}{\p{N-1}fU} \var_q\sp{\sin\p{qa}}
  \p{\xi_1^2 \D_{1,0}^2 - 2 N \xi_2 \D_{2,0}}
  + O\p{\phi^3},
  \label{eq:H_eff_2_phi_apndx}
\end{align}
where the factors $\xi_L$ are defined in \eqref{eq:scale_fac},
\begin{align}
  \var_q\sp{X_q} \equiv \cov_{p,q}\sp{X_p,X_q}
  = \EE_p\sp{\p{X_p-\EE_q\sp{X_q}}^2}
  = \EE_p\sp{X_p^2} - \EE_q\sp{X_q}^2
\end{align}
is the variance of $X_q$ over all occupied quasi-momenta $q$, and we
have neglected scalar terms $\sim\D_{0,0}\propto\1$ that have no
physical consequence.  Note that in the case of SU(2), the drive
operator $\D_{1,0}$ is precisely the collective spin-$z$ operator
$S_\z$ that appears in Ref.~\cite{he2019engineering}.  Furthermore,
drive operators $D_{LM}$ vanish when $L\ge n$, so in the case of SU(2)
the operator $\D_{2,0}$ has no contribution to the effective
Hamiltonians in \eqref{eq:H_eff_2_phi_apndx} and
\eqref{eq:H_eff_2_phi_apndx}.
